---
title: "Building and Managing Local Databases from Google Earth Engine using geeLite"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{geeLite}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include = FALSE}
pkgs <- c("dplyr", "leaflet", "withr", "sf")
if (length(pkgs <- setdiff(pkgs, rownames(installed.packages()))))
  install.packages(pkgs, repos = "https://cloud.r-project.org")
rm(pkgs)

# Define a temporary path for SQLite database
path_to_db <- withr::local_tempdir()
```

```{r setup}
# Load the necessary packages
library(geeLite)
library(withr)

# Set the time zone to UTC temporarily during vignette execution
withr::local_timezone("UTC")
```

# Introduction

`geeLite` simplifies the process of building, managing, and updating local SQLite databases containing geospatial features extracted from Google Earth Engine (GEE). This vignette covers the installation, configuration, data collection, and analysis workflows using the `geeLite` package.

# Installation

To install `geeLite` from GitHub, use the following commands:

```{r, eval = FALSE}
# Install devtools if not installed
# install.packages("devtools")

# Install geeLite from GitHub
devtools::install_github("mtkurbucz/geeLite")

# Install Python dependencies and setup rgee
geeLite::gee_install()
```

# Workflow

This section describes the workflow for setting up and working with `geeLite`, including configuration, data collection, modification, and data analysis.

## 1. Configuration

The first step in using `geeLite` is setting up a configuration file. This file specifies the regions of interest, the datasets to collect from GEE, and other parameters for data collection.

```{r, eval = FALSE, warning = FALSE}
# Define the path for the SQLite database
path <- path_to_db

# Create a configuration for Somalia (SO) and Yemen (YE) to collect NDVI data
set_config(
  path = path,
  regions = c("SO", "YE"),
  source = list(
    "MODIS/061/MOD13A2" = list(
      "NDVI" = c("mean", "sd")
    )
  ),
  resol = 3,
  start = "2020-01-01"
)
```

This configuration will create a JSON file at the specified path that defines the parameters for collecting data from GEE.

## 2. Data Collection

Once the configuration is set, you can collect data from GEE using the `run_geelite()` function. This function retrieves data based on the configuration file and stores it in a local SQLite database.

```{r, eval = FALSE}
# Collect the data and store it in the SQLite database
run_geelite(path = path)
```

The function will store the collected geospatial data in the SQLite database and log the progress.

## 3. Modifying the Configuration

If you want to modify the configuration (e.g., to add new statistics or datasets), you can use the `modify_config()` function to make updates without rebuilding the entire database.

```{r, eval = FALSE}
# Add more statistics to the NDVI band and include EVI data
modify_config(
  path = path,
  keys = list(
    c("source", "MODIS/061/MOD13A2", "NDVI"),
    c("source", "MODIS/061/MOD13A2", "EVI")
  ),
  new_values = list(
    c("mean", "min", "max"),
    c("mean", "sd")
  )
)
```

After modifying the configuration, run `run_geelite()` again to update the database with the new settings.

## 4. Reading and Analyzing Data

Once the data has been collected and stored in the database, you can read and analyze it using the `read_db()` function. This function allows you to aggregate data at different frequencies (e.g., daily, monthly) and apply preprocessing functions.

```{r, eval = FALSE}
# Read the data from the database, aggregate to monthly frequency by default
db <- read_db(path = path)

# Read the data with custom aggregation functions
db <- read_db(path = path, aggr_funs = list(
  function(x) mean(x, na.rm = TRUE),
  function(x) sd(x, na.rm = TRUE))
)
```

# Usage Example

Here is an example of how to use `geeLite` to collect NDVI data for Somalia and Yemen, aggregate it to a monthly frequency, and visualize it using the `leaflet` package.

## 1. Collecting the Data

```{r, warning = FALSE}
# Define the path for the SQLite database
path <- path_to_db

# Set the configuration file for NDVI data collection
set_config(
  path = path,
  regions = c("SO", "YE"),
  source = list(
    "MODIS/061/MOD13A2" = list(
      "NDVI" = c("mean", "sd")
    )
  ),
  resol = 3,
  start = "2020-01-01"
)

# Collect the data
run_geelite(path = path)

# Read the data from the database
db <- read_db(path = path, freq = "month")
```

## 2. Visualizing the Data

Once the data is collected, you can visualize it using the `leaflet` package. The following code shows how to plot the mean NDVI values for each region.

```{r}
# Load necessary packages
library(leaflet)
library(dplyr)
library(sf)

# Read database and merge grid with MODIS data
sf <- merge(db$grid, db$`MODIS/061/MOD13A2/NDVI/mean`, by = "id")

# Select the date to visualize
ndvi <- sf$`2020-01-01`

# Create a color palette function based on the values
color_pal <- colorNumeric(palette = "viridis", domain = ndvi)

# Create the leaflet map
leaflet(data = sf) %>%
  addTiles() %>%                            # Add base tiles
  addPolygons(
    fillColor = color_pal(ndvi),            # Fill color
    color = "#BDBDC3",                      # Border color
    weight = 1,                             # Border weight
    opacity = 1,                            # Border opacity
    fillOpacity = 0.9                       # Fill opacity
  ) %>%
  addScaleBar(position = "bottomleft") %>%  # Add scale bar
  addLegend(
    pal = color_pal,                        # Color palette
    values = ndvi,                          # Data values to map
    title = "Mean NDVI",                    # Legend title
    position = "bottomright"                # Legend position
  )
```

# Extras

In this section, we explore additional features of the `geeLite` package, such as using the command-line interface (CLI) for automation and integrating the package with job scheduling systems like `cron`.

## 1. Command-Line Interface (CLI)

`geeLite` provides a CLI that allows you to run the main functions of the package directly from the command line. This is useful for automating workflows or integrating `geeLite` into larger systems.

```{bash, eval = FALSE}
# Setting the CLI files
Rscript /path/to/geeLite/cli/set_cli.R --path "path/to/db"

# Set up the configuration via CLI
Rscript cli/set_config.R --regions "SO YE" --source "list('MODIS/061/MOD13A2' = list('NDVI' = c('mean', 'min')))" --resol 3 --start "2020-01-01"

# Collecting GEE data via CLI
Rscript cli/run_geelite.R

# Modifying the configuration via CLI
Rscript cli/modify_config.R --keys "list(c('source', 'MODIS/061/MOD13A2', 'NDVI'), c('source', 'MODIS/061/MOD13A2', 'EVI'))" --new_values "list(c('mean', 'min', 'max'), c('mean', 'sd'))"

# Updating the database via CLI
Rscript cli/run_geelite.R
```

## 2. Automating Data Collection with Cron

You can automate database updates using the Linux `cron` job scheduler. Here’s an example cron job that updates the database monthly:

```{bash, eval = FALSE}
# Open the cron jobs file
crontab -e

# Add the following line to schedule monthly updates (runs on the 1st of every month)
0 0 1 * * Rscript /path/to/db/cli/run_geelite.R
```

# Conclusion

The `geeLite` package simplifies the process of managing large-scale geospatial data from Google Earth Engine. With its easy-to-use functions and CLI support, you can efficiently build, update, and analyze local databases for geospatial analysis. By integrating with GEE and R's powerful data processing capabilities, `geeLite` is a valuable tool for researchers and analysts working with environmental and spatial data.

For more detailed information and updates, visit the [geeLite GitHub repository](https://github.com/mtkurbucz/geeLite/).

### Key Improvements:
1. **Workflow Structure**: Divided into clear steps such as *Configuration*, *Data Collection*, *Modification*, and *Reading and Analyzing Data*.
2. **Example Section**: Provided a concrete example of collecting, processing, and visualizing NDVI data.
3. **Extras Section**: Added sections on CLI usage and automation with `cron`.
4. **CLI and Automation**: Highlighted how to use the CLI for automating processes, making it practical for advanced users.

This structure ensures that users can easily follow the workflow and understand how to use `geeLite` effectively in both interactive and automated environments.
